Characteristic polynomials of p-adic matrices

### 2017-02-04 charpoly/charpoly.tex
Working on Intro

\todo{Explain here why computing characteristic polynomial is so important.}

@@

\todo{Explain also why taking care of precision is important.}

@@

A faster approach over a field is to compute the Frobenius normal form of $M$,
which is achievable in running time $\softO(n^\omega)$ \cite{storjohann:01a}.
However, the fact that it uses division frequently leads to catastrophic losses of division.
In many examples, no precision remains at the end of the calculation.

@@

\todo{Example of $M$ and $M + 1$.}

One benefit of this approach is that we may account for diffuse precision:
precision that is not localized on any single coefficient of $\chi_M$.  For example,
let $0 \le \alpha_1 \le \alpha_2 \le \dots \le \alpha_n$, consider a
diagonal matrix $D$ with diagonal entries $(p^{\alpha_1}, \dots, p^{\alpha_n})$,
let $P, Q \in \GL_n(\Zp)$ and set $M = PDQ$.  The valuation of the coefficient
of $X^{n-k}$ in $\chi_M$ will be $\sum_{i=1}^k \alpha_i$, and if $M$ is known
with precision $O(p^m)$ then $\chi_M$ will be known with some precision
larger than $O(p^m)$ (see Proposition \ref{prop:mainlem}).

@@

\subsection*{Notation} Throughout the paper, $K$ will refer to a complete,
discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
$\OK$ its ring of integers and $\pi$ a uniformizer. \todo{Soft-Oh}

@@

\subsection{The theory of p-adic precision}

@@

\begin{cor}
Suppose that $M \in \GL_n(\OK)$ is known with precision $O(p^m)$.
Then the characteristic polynomial of $M$ has precision larger than $O(p^m)$
if and only if the reduction of $M$ modulo $\pi$ does not have a cyclic vector.
\end{cor}

Note that this criterion is checkable using $\softO(n^\omega)$ operations in the residue field,
the running time of matrix multiplication.

\todo{Image modulo $p$ and elementary divisors. 
Consequences: (1)~computation in $\softO(n^\omega)$ and (2)~gain criterium.}

@@

Proposition~\ref{prop:shortcom} shows that $\com(X{-}M)$ can be encoded 
by the datum of the quadruple $(\alpha, P, Q, \chi_M)$ whose total size 
stays within $O(n^2)$ : the polynomials $\alpha$ and $\chi_M$ are 
determined by $2n$ coefficients while we need $2n^2$ coefficients to 
write down the matrices $P$ and $Q$. 
We shall see moreover in \S \ref{sec:eigenvalues} that interesting
informations can be read off on this short form $(\alpha, P, Q, 
\chi_M)$.

@@

\begin{rem} \label{rem:char_pol_from_hessenberg}
From an Hessenberg form of $M,$ it is well known
that one can compute the characteristic polynomial of 
$M$ in $O(n^3)$ operations in $K.$ See \cite{Cohen:2013} page
55 and 56.
However, this computation involves some divisions, and its
behaviour regarding to precision is not easy to
quantify.
\end{rem}

@@

As a consequence of the previous discussion, once the optimal jagged 
precision is known, it is possible to lift arbitrarily the precision on 
$M$ and then to perform a computation of the characteristic polynomial 
at the optimal precision. This requires $\softO(n^3)$ operations in $K$ 
but, for several instances, may require to increase a lot the precision.

>>>>

@@

@@

A faster approach over a field is to compute the Frobenius normal form of $M$,
which is achievable in running time $\softO(n^\omega)$ \cite{storjohann:01a}.
However, the fact that it uses division frequently leads to catastrophic losses of precision.
In many examples, no precision remains at the end of the calculation.

@@

One benefit of this approach is that we may account for diffuse precision:
precision that is not localized on any single coefficient of $\chi_M$.  For example,
let $0 \le \alpha_1 \le \alpha_2 \le \dots \le \alpha_n$, consider a
diagonal matrix $D$ with diagonal entries $(p^{\alpha_1}, \dots, p^{\alpha_n})$,
let $P, Q \in \GL_n(\Zp)$ and set $M = PDQ$.  The valuation of the coefficient
of $X^{n-k}$ in $\chi_M$ will be $\sum_{i=1}^k \alpha_i$, and if $\alpha_{n-1} > 0$
and $M$ is known with precision $O(p^m)$ then the constant term of $\chi_M$
will be known with precision larger than $O(p^m)$ (see
\cite{caruso-roe-vaccon:15a}*{Prop. 3.2}).

As long as none of the eigenvalues of $M$ are congruent to $-1$ modulo $p$,
then none of the coefficients of the characteristic polynomial of $1+M$ will have
precision larger than $O(p^m)$.  But $\chi_{1+M}(X) = \chi_M(X-1)$, so the precision
content of these two polynomials should be equivalent. The solution is that the
extra precision in $\chi_{1+M}$ is diffuse and not visible on any individual coefficient.
We formalize this phenomenon using lattices; see Section \ref{sec:padicprec} for
further explanation, and \cite{caruso:17a}*{\S 3.2.2} for a specific example of the
relationship between $\chi_M$ and $\chi_{1+M}$.

@@

\subsection*{Notation} Throughout the paper, $K$ will refer to a complete,
discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
$\OK$ its ring of integers and $\pi$ a uniformizer. \todo{$\softO$, $\omega$}

@@

\subsection{The theory of p-adic precision} \label{sec:padicprec}

@@

\begin{cor} \label{cor:prec_gain}
Suppose that $M \in \GL_n(\OK)$ is known with precision $O(p^m)$.
Then the characteristic polynomial of $M$ has precision lattice strictly contained in $O(p^m)$
if and only if the reduction of $M$ modulo $\pi$ does not have a cyclic vector.
\end{cor}

Note that this criterion is checkable using $\softO(n^\omega)$ operations in the residue field \cite{storjohann:01a}.

@@

Proposition~\ref{prop:shortcom} shows that $\com(X{-}M)$ can be encoded 
by the datum of the quadruple $(\alpha, P, Q, \chi_M)$ whose total size 
stays within $O(n^2)$ : the polynomials $\alpha$ and $\chi_M$ are 
determined by $2n$ coefficients while we need $2n^2$ coefficients to 
write down the matrices $P$ and $Q$. 
We shall see moreover in \S \ref{sec:diffFrob} that interesting
informations can be read off on this short form $(\alpha, P, Q, 
\chi_M)$.

@@

\begin{rem} \label{rem:char_pol_from_hessenberg}
From a Hessenberg form of $M,$ it is well known
that one can compute the characteristic polynomial of 
$M$ in $O(n^3)$ operations in $K$ \cite{Cohen:2013}*{pp. 55--56}.
However, this computation involves division, and its
precision behavior is not easy to quantify.
\end{rem}

@@

\begin{rem}
As a consequence of the previous discussion, once the optimal jagged 
precision is known, it is possible to lift the entries of $M$ to a sufficiently
large precision and then use 
and then to perform a computation of the characteristic polynomial 
at the optimal precision. This requires $\softO(n^3)$ operations in $K$ 
but, for several instances, may require to increase a lot the precision.
\end{rem}
